%% swarm-abm — manuscript for Simulation Modelling Practice and Theory (SIMPAT)
%% Elsevier elsarticle class. Submission format: [review] (single column,
%% double spaced, line numbers). Final two-column look: replace with [5p].
\documentclass[review]{elsarticle}

\usepackage{lineno}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\modulolinenumbers[5]

\journal{Simulation Modelling Practice and Theory}
\bibliographystyle{elsarticle-num}

\begin{document}

\begin{frontmatter}

\title{swarm-abm: a deterministic, multi-paradigm engine for spatial
agent-based simulation with compiler-verified parallel updates}

%% TODO: author names, affiliations, corresponding author
\author[a]{First Author\corref{cor1}}
\ead{author@example.org}
\cortext[cor1]{Corresponding author}
\affiliation[a]{organization={Affiliation},
            city={City},
            country={Country}}

\begin{abstract}
Agent-based modelling (ABM) underpins research across ecology, epidemiology,
the social sciences and the geosciences, yet the reference platforms ---
NetLogo and Mesa --- trade performance for accessibility, and their per-run
cost constrains experimental design. We present swarm-abm, a spatial ABM
engine in Rust that unifies three spatial paradigms (regular grid, network and
continuous space) under a single agent/model API, with bit-for-bit
deterministic, seedable random number generation. Its core methodological
contribution is a two-phase synchronous (simultaneous) update whose read-only
decision phase is enforced by the type system: the compiler proves that no
agent can mutate shared state during decision, which (i) removes a classic
verification pitfall of synchronous ABM --- read-after-write --- by
construction, and (ii) makes the decision phase safe to parallelise. Combined
with a per-agent deterministic generator, parallel execution is bit-for-bit
identical to sequential regardless of thread count; we validate the generator
stream with PractRand and confirm identical results across CPU (x86-64) and
WebAssembly (wasm32) targets. On two canonical models (spatial SIR, Schelling),
swarm-abm runs 2--5$\times$ faster single-threaded than Agents.jl --- the
fastest compiled ABM framework --- and 45--184$\times$ faster than Mesa; the
parallel decision phase scales $\sim$5$\times$ on 16 cores for compute-bound
decisions. We demonstrate the engine on a real application whose performance
enables a statistically powered comparison of five metaheuristic calibrators,
and provide Python bindings and a WebAssembly viewer. The engine is open
source.
\end{abstract}

\begin{keyword}
agent-based modelling \sep simulation framework \sep reproducibility \sep
parallel simulation \sep verification and validation \sep Rust
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}
\label{sec:intro}

Agent-based modelling (ABM) is a workhorse method across ecology, epidemiology,
the social sciences and the geosciences: it builds system-level behaviour from
the bottom up, out of local interactions among many heterogeneous agents. Two
platforms dominate practice --- NetLogo~\cite{wilensky1999netlogo}, on the Java
virtual machine, and Mesa~\cite{kazil2020mesa}, in Python. Both prioritise
accessibility, and both pay for it in performance.

Performance is not a mere convenience here: \emph{the cost of a single run
shapes the science that gets done}. When one replicate of a spatial model takes
minutes, studies default to a single run per condition, calibrations are tuned
to a single random seed (and silently overfit it), and sensitivity analyses are
truncated. The methodological consequences are real, and they are invisible in
the published result. A faster engine does not merely save wall-clock time; it
widens the space of experimental designs that are feasible at all.

We identify three gaps in current spatial ABM tooling, and a fourth concern
that is specific to \emph{synchronous} (simultaneous) update. (1)
\textbf{Reproducibility}: many models rely on a global, unseeded generator, so
results cannot be replicated bit-for-bit --- a weak foundation for a
computational science. (2) \textbf{Performance}: the per-run cost forces the
impoverished experimental designs noted above. (3) \textbf{Spatial generality}:
tools tend to privilege one spatial paradigm, so moving a question from a grid
to a network to a continuous space typically means a rewrite. (4) The
\textbf{synchronous-update pitfall}: in two-phase (decide/apply) update, every
agent must decide against the \emph{same} snapshot of the world; if any agent
reads a neighbour's freshly written state during the decision phase
(read-after-write), the update is silently corrupted. In existing frameworks
this is a matter of user discipline, not a guarantee.

This paper presents \textbf{swarm-abm}, a spatial ABM engine written in Rust
that addresses all four. Its contributions are:

\begin{itemize}
  \item A \textbf{multi-paradigm engine} that exposes a regular grid, a network
  (graph) and a continuous space under one \texttt{Agent}/\texttt{Model} API,
  with seedable, portable, bit-for-bit deterministic random numbers
  (Section~\ref{sec:design}).
  \item A \textbf{compiler-verified synchronous update}: the decision phase
  receives the model immutably, so the type system \emph{proves} that no agent
  writes shared state during decision. This removes read-after-write by
  construction (a verification guarantee, not a convention) and is precisely
  what makes the phase safe to parallelise (Section~\ref{sec:sync}).
  \item \textbf{Deterministic parallelism}: a per-agent generator keyed on
  $(\text{seed}, \text{step}, \text{agent})$ makes parallel execution
  bit-for-bit identical to sequential, independent of thread count. We validate
  the generator with PractRand and confirm bit-identical results across x86-64
  and WebAssembly (Section~\ref{sec:vv}).
  \item A \textbf{performance evaluation against the real peers}: on spatial SIR
  and Schelling, swarm-abm is 2--5$\times$ faster single-threaded than
  Agents.jl~\cite{datseris2022agents} --- the fastest compiled ABM framework ---
  and 45--184$\times$ faster than Mesa; the parallel decision phase scales
  $\sim$5$\times$ on 16 cores when per-agent decisions are compute-bound
  (Section~\ref{sec:perf}).
  \item A \textbf{real-world demonstration} in which the engine's throughput
  turns a previously infeasible study --- a statistically powered comparison of
  five metaheuristic calibrators --- into a routine one
  (Section~\ref{sec:app}), plus Python bindings and a WebAssembly viewer
  (Section~\ref{sec:access}).
\end{itemize}

The remainder of the paper is organised as follows.
Section~\ref{sec:design} describes the engine.
Section~\ref{sec:sync} develops the synchronous-update guarantee and
deterministic parallelism. Section~\ref{sec:vv} reports verification and
validation. Section~\ref{sec:perf} reports performance and scalability.
Section~\ref{sec:app} presents the application. Section~\ref{sec:access} covers
accessibility and reproducibility. Sections~\ref{sec:discussion}
and~\ref{sec:conclusion} discuss and conclude.

\section{Engine design}
\label{sec:design}
% TODO: traits Agent/Model; take-out borrow pattern; tres espacios (Grid2D,
% Graph, ContinuousSpace) bajo una API; scheduler (ordered/random/simultaneous);
% SimRng ChaCha8; DataCollector; batch (run_ensemble/run_sweep, rayon). Fig 1.

\section{Synchronous update as a verification guarantee, and deterministic parallelism}
\label{sec:sync}
% TODO: la fase decide recibe &Model inmutable -> el compilador prueba no-escritura;
% read-after-write imposible por construccion. Doble buffer via mem::take.
% RNG por-agente child_rng(seed,step,id) -> paralelo == secuencial bit a bit.
% Teorema/propiedad + referencia al test parallel_decide.rs.

\section{Verification and validation}
\label{sec:vv}
% TODO: (a) paridad distribucional vs Mesa (7/7 metricas, |z|<=1.22, curvas);
% (b) RNG: PractRand sin anomalias 1GB en stream inter-agente; (c) determinismo
% cross-platform x86-64 vs wasm32 bit-identico (tabla hex).

\section{Performance and scalability}
\label{sec:perf}
% TODO: cross-engine SIR+Schelling vs Agents.jl y Mesa (tabla ms/paso, speedups);
% metodologia (warmup JIT, build fuera del cronometro, mediana 3 seeds).
% Escalabilidad del decide paralelo: compute-bound ~5.3x/16 hilos, memory-bound
% ~1.6x; Amdahl (apply secuencial). Fig speedup.

\section{Application: spatial debris-flow simulation}
\label{sec:app}
% TODO: port fiel de un modelo Mesa/Python (Atacama 2015), paridad distribucional,
% ~100x. Enmarcar como reproduccion (citar manuscrito ESPL en revision, SIN las
% mejoras). El aporte M&S: el rendimiento habilita (i) calibracion robusta
% multi-semilla; (ii) benchmark de 5 metaheuristicas (DE/GA/PSO/SA/GWO) con
% Friedman + Wilcoxon-Holm \cite{demsar2006statistical} = procedimiento de
% seleccion/comparacion. Fig.

\section{Accessibility and reproducibility}
\label{sec:access}
% TODO: bindings Python (PyO3) modelos nativos + barridos paralelos; visor WASM;
% open source MIT/Apache; CI; deposito archivado (Zenodo DOI, TODO).

\section{Discussion}
\label{sec:discussion}
% TODO: compute-no-es-neutral (decisiones cientificas habilitadas);
% determinismo como requisito metodologico; generalidad espacial; comparacion
% honesta con Agents.jl/Repast HPC \cite{collier2013repast}/FLAME GPU
% \cite{richmond2023flamegpu}. Limitaciones: definir modelos requiere Rust
% (mitigado por bindings); apply aun secuencial (techo de Amdahl).

\section{Conclusions}
\label{sec:conclusion}
% TODO: motor determinista, multi-paradigma, de alto rendimiento, validado vs
% Mesa y Agents.jl, con garantia de V&V para update sincrono y paralelismo
% reproducible; demostrado en un caso real. Amplia el espacio de analisis tratables.

\section*{Declaration of generative AI use}
% TODO (obligatorio Elsevier si aplica).

\bibliography{refs}

\end{document}
